D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-09 ootavacuous.tex
Clean up UB-blocking discussion
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Peter Sewell feedback, initial installment
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	This paper goes further, adding temporal constraints from
	long-standing laws of physics and constraints on hardware
	architecture and design to demonstrate that OOTA cannot occur in
	correctly constructed C++ programs and correctly constructed C++
	implementations running on on real-world systems.
	This paper also looks at some tooling implicatoins of OOTA.
\end{abstract}

@@

This section provides a brief overview of the OOTA problem followed
by an equally brief summary of prior work in this area.

@@

This \co{mightnotreturn} clobber is merely an example.
There might well be better ways to prevent UB from propagating backwards.
Other options being considered include a call to an empty function that
is invisible to the compiler and a GCC goto asm that the compiler cannot
prove falls through.
@@@ show details.

A \co{std::observable()} function has been proposed to the C++ standards
committee~\cite{DavisHerring2021P1494R2}.

@@

% @@@ reviewer to do.
% Luc Maranget: Memory models.
% Richard Gristenthwaite: ARM hardware.
% Mark Rutland: ARM memory ordering.
% Will Deacon: ARM hardware, C++, and OOTA.
% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
% Christoph Hellwig: Why slow on BPF memory model.
% Miguel Ojeda: LKMM and Rust.
% Alice Ryhl: LKMM and Rust.
% Ralf Jung: Rust.
% Catalin Marinas: ARM and formal methods (qspinlock).
% Segher Boessenkool:  GCC.
% Dan Lustig: NVIDIA memory ordering.
% Andrea Parri: Memory models, RISC-V hardware.
% Jonas Oberhauser: Memory models.
% Hernan Ponce de Leon: Memory models.
% Palmer Dabbelt: RISC-V hardware.
% Ali Sezgin: P0422R0.
% Tony Tye: P0422R0.
% Other memory-model maintainers, including Peter Zijlstra.
% Derek Williams: PowerPC hardware.
% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
% Olivier Giroux: Skeptic, memory models, NVIDIA.
% Luke Geeson: Whatever...  Review.
% Ori Lahav: Prior OOTA.
% Viktor Vafeiadis:  Prior OOTA.
% Derek Dreyer: Prior OOTA.
% Brian Demsky: Prior OOTA.

@@

% Done as of November 30, 2023:
% David Goldblatt
% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
%	Frederic Weisbecker (RCU proteges)
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.

% @@@ TODO:
% Add more detail on blocking UB to "Aside on Undefined Behavior".

>>>>

	This paper adds constraints from long-standing laws of physics
	and on hardware systems to demonstrate that OOTA cannot occur
	in correctly constructed C++ programs running on correctly
	constructed C++ implementations.

	In this, OOTA behaviors are similar to the issues surrounding
	undecidability and its real-world counterpart, high complexity.
	So similar, in fact, that not only the same strategies apply
	to OOTA, but also the same \emph{code} applies, both for C++
	implementations and for associated tooling.
\end{abstract}

@@

This section provides a brief overview of the OOTA problem, followed
by an equally brief summary of prior work in this area, and ending
with a quick review of a few of the challenges for general definitions of
``OOTA cycle''.

@@

This \co{mightnotreturn} clobber is merely an example.
There might well be better ways to prevent UB from propagating backwards.
Other options being considered include a call to an empty function that
is invisible to the compiler and a GCC goto asm that the compiler cannot
prove falls through.

A \co{std::observable()} function that would block backwards-propagating
UB has been proposed to the C++ standards
committee~\cite{DavisHerring2021P1494R2}.

@@

% @@@ reviewer to do.
% Luc Maranget: Memory models.
% Mark Rutland: ARM memory ordering.
% Will Deacon: ARM hardware, C++, and OOTA.
% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
% Christoph Hellwig: Why slow on BPF memory model.
% Miguel Ojeda: LKMM and Rust.
% Alice Ryhl: LKMM and Rust.
% Ralf Jung: Rust.
% Catalin Marinas: ARM and formal methods (qspinlock).
% Segher Boessenkool:  GCC.
% Dan Lustig: NVIDIA memory ordering.
% Andrea Parri: Memory models, RISC-V hardware.
% Jonas Oberhauser: Memory models.
% Hernan Ponce de Leon: Memory models.
% Palmer Dabbelt: RISC-V hardware.
% Ali Sezgin: P0422R0.
% Tony Tye: P0422R0.
% Other memory-model maintainers, including Peter Zijlstra.
% Derek Williams: PowerPC hardware.
% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
% Olivier Giroux: Skeptic, memory models, NVIDIA.
% Luke Geeson: Whatever...  Review.
% Ori Lahav: Prior OOTA.
% Viktor Vafeiadis:  Prior OOTA.
% Derek Dreyer: Prior OOTA.
% Brian Demsky: Prior OOTA.

@@

% Done as of November 30, 2023:
% David Goldblatt
% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
%	Frederic Weisbecker (RCU proteges)
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.
% Martin Uecker (offered)
% Peter Sewell
% Mark Batty
% Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.

% @@@ TODO:
% Measure store-buffer delays on an x86 system.
